---
title: "Project by Group Validation"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Validation Group Project

## Grading Criteria

**Framing & data (15%)**

-   Clear question + time unit + scope − Data suits time-series (enough
    history, sensible frequency)

**Exploratory time-series insights (20%)**

-   Identifies trend/seasonality/breaks/variance − Uses appropriate
    visuals (plots/ACF/PACF/decomp) − Interpretation links to modelling
    choices

**Modelling choices & justification (25%)**

-   Benchmarks + at least one ETS/ARIMA/SARIMA (as appropriate) − Clear
    justification (transform/difference/seasonality) − Tries
    alternatives (not single “auto” only)

**Forecast evaluation on future data (20%)**

-   Time-based train/test split (or rolling) − Uses metrics (e.g.,
    MAE/RMSE/MASE) + brief interpretation − Forecast vs actual on test
    window; comparison drives selection

**Diagnostics & conclusions (15%)**

-   Residual plot + ACF + Ljung–Box/portmanteau − States whether
    residuals look like white noise − Clear conclusion +
    limitations/assumptions

## Project

### 1. Data & question

We use daily Brent crude oil prices as our time series dataset. The question is: Can we accurately forecast Brent oil prices using classical time series models (ETS, ARIMA, SARIMA)?

- Data source: U.S. Energy Information Administration (EIA) or FRED (Federal Reserve Economic Data)
- Time unit: Daily
- Scope: Multiple years (e.g., 2010–2025)
- The dataset contains historical daily closing prices for Brent crude oil, suitable for time series analysis with sufficient history and frequency.


```{r libraries}
library(fpp3)
library(tsibbledata)
library(tidyverse)
library(lubridate)
library(quantmod)
library(dygraphs)
library(forecast)
library(prophet)
library(quantmod)
```


```{r oil price data}
# Get WTI crude oil futures (symbol: CL=F) from Yahoo Finance
library(quantmod)
getSymbols("CL=F", src = "yahoo", from = "2010-01-01", auto.assign = TRUE)
wti <- `CL=F`

# Convert to tibble and preprocess
library(tidyverse)
library(tsibble)
wti_tbl <- as_tibble(wti) %>%
  mutate(date = index(wti)) %>%
  select(date, price = `CL=F.Close`) %>%
  filter(!is.na(price)) %>%
  arrange(date) %>%
  as_tsibble(index = date)

glimpse(wti_tbl)
summary(wti_tbl$price)
```

```{r load-electricity-data}
# Load Ember European Wholesale Electricity Price Data
library(readr)
electricity_data <- read_csv("european_wholesale_electricity_price_data_monthly.csv")
glimpse(electricity_data)
summary(electricity_data$`Price (EUR/MWhe)`)
```
```

### 2. Initial visual inspection

```{r wti-visual-inspection, message=FALSE, warning=FALSE}
# Get WTI crude oil futures (symbol: CL=F) from Yahoo Finance
library(quantmod)
getSymbols("CL=F", src = "yahoo", from = "2010-01-01", auto.assign = TRUE)
wti <- `CL=F`

# Convert to tibble and preprocess
library(tidyverse)
library(tsibble)
wti_tbl <- as_tibble(wti) %>%
  mutate(date = index(wti)) %>%
  select(date, price = `CL=F.Close`) %>%
  filter(!is.na(price)) %>%
  arrange(date) %>%
  as_tsibble(index = date)

glimpse(wti_tbl)
summary(wti_tbl$price)

# Plot the time series
ggplot(wti_tbl, aes(x = date, y = price)) +
  geom_line(color = "steelblue") +
  labs(title = "WTI Crude Oil Futures Prices (2010–2025)",
       x = "Date",
       y = "Price (USD/barrel)") +
  theme_minimal()

# ACF and PACF plots
acf(wti_tbl$price, main = "ACF: WTI Oil Futures Prices")
pacf(wti_tbl$price, main = "PACF: WTI Oil Futures Prices")
```

Interpretation:
- The plot shows the trend and volatility in WTI oil futures prices over time.
- ACF/PACF plots help identify autocorrelation and potential seasonality or breaks.

### 3. Pre-processing choices (justify)

### 4. Baseline forecasts

### 5. Model Testing

-   ETS (exponential smoothing)

-   ARIMA

-   SARIMA

### 6. Evaluation on future data

-   Use a time-based train/test split (not random)

-   Compare models (e.g., MAE + RMSE, or MASE + RMSE)

-   Show a plot of forecasts vs actuals on the test period.

### 7. Residual diagnostics

### 8. Conclusion
