---
title: "Project by Group Validation"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Validation Group Project

## Grading Criteria

**Framing & data (15%)**

-   Clear question + time unit + scope − Data suits time-series (enough
    history, sensible frequency)

**Exploratory time-series insights (20%)**

-   Identifies trend/seasonality/breaks/variance − Uses appropriate
    visuals (plots/ACF/PACF/decomp) − Interpretation links to modelling
    choices

**Modelling choices & justification (25%)**

-   Benchmarks + at least one ETS/ARIMA/SARIMA (as appropriate) − Clear
    justification (transform/difference/seasonality) − Tries
    alternatives (not single “auto” only)

**Forecast evaluation on future data (20%)**

-   Time-based train/test split (or rolling) − Uses metrics (e.g.,
    MAE/RMSE/MASE) + brief interpretation − Forecast vs actual on test
    window; comparison drives selection

**Diagnostics & conclusions (15%)**

-   Residual plot + ACF + Ljung–Box/portmanteau − States whether
    residuals look like white noise − Clear conclusion +
    limitations/assumptions

## Project

### 1. Data & question

In the last years, Europe has faced geopolitical hardships that have led to significant volatility in energy prices. A large part of this has had to do with gas and oil dependency on other countries, in special Russia. More recently, the conflict in the Gulf has led to gas supplies being cut off from the European market, which has led to a surge in gas prices once again. This has had a significant impact on energy prices in Germany, which is one of the largest consumers of energy in Europe.
This begs the question: to what extent do gas or oil prices affect energy prices in Germany? More specifically, after controlling for one of these commodities, can we better predict energy prices?

```{r libraries}
library(fpp3)
library(tsibbledata)
library(tidyverse)
library(lubridate)
library(quantmod)
library(dygraphs)
library(forecast)
library(prophet)
library(quantmod)
library(readr)
```


```{r load germany electricity data}
# Load Ember European Wholesale Electricity Price Data
electricity_data <- read_csv("european_wholesale_electricity_price_data_monthly.csv")

german_electricity_data <- electricity_data %>%
  filter(Country == "Germany") %>%
  mutate(date = ymd(Date)) %>%
  mutate(price = `Price (EUR/MWhe)`) %>%
  select(date, price)

glimpse(german_electricity_data)
summary(german_electricity_data$price)
```

```{r load european gas prices}
# Get TTF Natural Gas (symbol: TTF=F) from Yahoo Finance
getSymbols("TTF=F", src = "yahoo", from = "2015-01-01", auto.assign = TRUE)
gas <- `TTF=F`

gas_tbl <- as_tibble(gas) %>%
  mutate(date = index(gas)) %>%
  select(date, price = `TTF=F.Close`) %>%
  arrange(date) %>%
  as_tsibble(index = date)

glimpse(gas_tbl)
```

```{r load brent futures data}
# Get Brent crude oil futures (symbol: BZ=F) from Yahoo Finance
getSymbols("BZ=F", src = "yahoo", from = "2015-01-01", auto.assign = TRUE)
brent <- `BZ=F`

# Convert to tibble and preprocess
brent_tbl <- as_tibble(brent) %>%
  mutate(date = index(brent)) %>%
  select(date, price = `BZ=F.Close`) %>%
  arrange(date) %>%
  as_tsibble(index = date)

glimpse(brent_tbl)
```

### 2. Initial visual inspection

```{r german electricity visual inspection}
ggplot(german_electricity_data, aes(x = date, y = price)) +
  geom_line(color = "steelblue") +
  labs(title = "German Wholesale Electricity Prices",
       x = "Date",
       y = "Price (EUR/MWhe)") +
  theme_minimal()
```

```{r indexed energy vs oil prices, echo=TRUE}
# Index all series from their first value
energy_base <- german_electricity_data$price[german_electricity_data$date == min(german_electricity_data$date)]
brent_base <- brent_tbl$price[brent_tbl$date == min(brent_tbl$date)]
gas_base <- gas_tbl$price[gas_tbl$date == min(gas_tbl$date)]

energy_data_indexed <- german_electricity_data %>% mutate(indexed_price = (price / energy_base) * 100, series = "Energy Price (Germany)")
brent_tbl_indexed <- brent_tbl %>% mutate(indexed_price = (price / brent_base) * 100, series = "Brent Oil Price")
gas_tbl_indexed <- gas_tbl %>% mutate(indexed_price = (price / gas_base) * 100, series = "TTF Gas Price (JKM)")

plot_data <- bind_rows(energy_data_indexed %>% select(date, indexed_price, series),
                      brent_tbl_indexed %>% select(date, indexed_price, series),
                      gas_tbl_indexed %>% select(date, indexed_price, series))

# Plot
ggplot(plot_data, aes(x = date, y = indexed_price, color = series)) +
  geom_line() +
  labs(title = "Indexed Energy, Brent Oil, and LNG Prices (Base 100)",
       y = "Indexed Price (Base = 100)",
       color = "Legend") +
  theme_minimal()
```


There is a clear strong positive relationship between all three, though in special gas and energy.
We can thus analyze the predictability of energy prices using historical data, as well as  wehn controlling for gas prices.

### 3. Pre-processing choices (justify)
```{r acf and pacf inspection}
# ACF and PACF plots
acf(german_electricity_data$price, main = "ACF: German Electricity Prices")
pacf(german_electricity_data$price, main = "PACF: German Electricity Prices")
```

### 4. Baseline forecasts


### 5. Model Testing

-   ETS (exponential smoothing)

-   ARIMA

-   SARIMA

### 6. Evaluation on future data

-   Use a time-based train/test split (not random)

-   Compare models (e.g., MAE + RMSE, or MASE + RMSE)

-   Show a plot of forecasts vs actuals on the test period.

### 7. Residual diagnostics

### 8. Conclusion
