---
title: "Time Series Analysis: Monthly Airline Passenger Numbers (1949–1960)"
author: "Group [Your Names Here]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5
)
```

```{r libraries}
# Core time-series packages
library(fpp3)       # tsibble, fable, feasts — modern tidy time-series workflow
library(tsibble)    # tidy time-series data structure
library(feasts)     # feature extraction and statistics (ACF, PACF, decomposition)
library(fable)      # forecasting models (ETS, ARIMA, benchmarks)

# Utilities
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
```

---

# Data & Question {#data-question}

## Dataset

We use the classic **`AirPassengers`** dataset, which records the **monthly number of international airline passengers** (in thousands) from **January 1949 to December 1960**.  
This dataset is built into R and is widely used as a benchmark for time-series methods.

**Time unit:** monthly observations  
**Length:** 144 observations (12 years × 12 months)  
**Variable:** total passengers (thousands)

## Research Question

> *Can we accurately forecast monthly international airline passenger numbers for 1959–1960 (the final two years), using data from 1949 to 1958?*

More broadly, we want to understand the **trend**, **seasonal pattern**, and **variance behaviour** of the series, choose an appropriate model, and evaluate its out-of-sample forecast accuracy.

```{r load-data}
# Convert the built-in ts object to a tsibble for a tidy workflow
passengers_ts <- as_tsibble(AirPassengers) %>%
  rename(passengers = value,
         month      = index)

# Quick look at the data structure
glimpse(passengers_ts)
kable(head(passengers_ts, 12), caption = "First 12 observations") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Initial Visual Inspection {#visual-inspection}

## Raw Series Plot

```{r raw-plot}
passengers_ts %>%
  autoplot(passengers) +
  labs(
    title    = "Monthly International Airline Passengers (1949–1960)",
    subtitle = "Source: Box & Jenkins (1976) via R datasets",
    x        = "Month",
    y        = "Passengers (thousands)"
  ) +
  theme_bw()
```

**Observations:**

- **Upward trend:** passenger numbers grow steadily over the 12-year period, roughly doubling from ~120k to ~450k.  
- **Strong annual seasonality:** a clear recurring peak each summer (July–August) and a trough in winter (November–February).  
- **Increasing variance:** the amplitude of seasonal swings grows with the level — characteristic of *multiplicative* seasonality. This motivates a log transformation.  
- **No obvious structural breaks or outliers** within the series.

## ACF and PACF

```{r acf-pacf}
passengers_ts %>%
  gg_tsdisplay(passengers, plot_type = "partial", lag_max = 48) +
  labs(title = "ACF and PACF — Raw Passengers Series")
```

**Interpretation:**

- The ACF decays very slowly (long memory), confirming a strong **trend** component.  
- Spikes at lags 12, 24, 36 in the ACF confirm **annual seasonality**.  
- The PACF shows a dominant spike at lag 1, consistent with an AR(1) or integrated process.  
- These patterns suggest that differencing (both regular and seasonal) will be needed before fitting ARIMA models.

## STL Decomposition

```{r stl-decomp}
passengers_ts %>%
  model(STL(passengers ~ trend(window = 13) + season(window = "periodic"),
            robust = TRUE)) %>%
  components() %>%
  autoplot() +
  labs(title = "STL Decomposition of Passengers Series") +
  theme_bw()
```

**Interpretation:**

- The **trend** component rises continuously with no obvious change point.  
- The **seasonal** component shows a stable summer peak / winter trough; choosing `season(window = "periodic")` fixes this shape over time — justified because the pattern looks very consistent.  
- The **remainder** is small relative to trend and season, suggesting the decomposition captures most variation.  
- **Important:** the growing seasonal amplitude in the raw series is not fully handled by an additive decomposition — further confirmation that a log transform is warranted.

---

# Pre-processing Choices {#preprocessing}

## Log Transformation

The increasing variance (proportional to the level) is the classic signature of a **multiplicative** structure. Taking logarithms converts multiplicative relationships to additive ones, stabilising the variance.

```{r log-transform}
passengers_ts <- passengers_ts %>%
  mutate(log_passengers = log(passengers))

passengers_ts %>%
  autoplot(log_passengers) +
  labs(
    title    = "Log-Transformed Monthly Passengers",
    x        = "Month",
    y        = "log(Passengers)"
  ) +
  theme_bw()
```

After the log transform the seasonal amplitude is visually constant, confirming this was the right choice.

## Differencing

We check the number of regular and seasonal differences required using unit-root tests:

```{r differencing-tests}
# Kwiatkowski–Phillips–Schmidt–Shin (KPSS) test for regular differencing
passengers_ts %>% features(log_passengers, unitroot_kpss)

# Number of regular differences suggested
passengers_ts %>% features(log_passengers, unitroot_ndiffs)

# Number of seasonal differences suggested
passengers_ts %>% features(log_passengers, unitroot_nsdiffs)
```

```{r differenced-series}
passengers_ts <- passengers_ts %>%
  mutate(
    diff1_log = difference(log_passengers, 1),        # first difference
    diff1_12_log = difference(diff1_log, 12)          # seasonal difference on top
  )

# Inspect stationarity after both differences
passengers_ts %>%
  filter(!is.na(diff1_12_log)) %>%
  autoplot(diff1_12_log) +
  labs(
    title = "Doubly-Differenced Log Passengers (d=1, D=1, s=12)",
    x     = "Month",
    y     = "Δ₁Δ₁₂ log(Passengers)"
  ) +
  theme_bw()
```

**Justification:**

- The KPSS test rejects stationarity for the raw log series → at least one regular difference needed.
- `unitroot_nsdiffs()` suggests one seasonal difference (D = 1) because the series has a seasonal unit root.
- After both differences the series fluctuates around zero with roughly constant variance — consistent with stationarity.
- We will pass `d = 1, D = 1` explicitly to ARIMA models (or let `auto_arima` confirm this).

---

# Baseline Forecasts {#baselines}

We split the data: **train = 1949–1958 (120 obs)**, **test = 1959–1960 (24 obs)**.

```{r train-test-split}
train <- passengers_ts %>% filter(year(month) <= 1958)
test  <- passengers_ts %>% filter(year(month) >= 1959)

cat("Training observations:", nrow(train), "\n")
cat("Test observations:    ", nrow(test),  "\n")
```

```{r baseline-models}
# Fit all four standard benchmark models on the log scale
baseline_fits <- train %>%
  model(
    Mean    = MEAN(log_passengers),
    Naive   = NAIVE(log_passengers),
    Drift   = RW(log_passengers ~ drift()),
    SNaive  = SNAIVE(log_passengers ~ lag("year"))
  )

baseline_fits
```

```{r baseline-forecasts}
# Forecast over the 24-month test horizon
baseline_fc <- baseline_fits %>%
  forecast(h = 24)

# Plot on original (exponentiated) scale for interpretability
baseline_fc %>%
  autoplot(passengers_ts %>% filter(year(month) >= 1957),
           level = NULL) +
  labs(
    title  = "Baseline Forecasts vs Actuals (1957–1960)",
    x      = "Month",
    y      = "log(Passengers)",
    colour = "Model"
  ) +
  theme_bw()
```

**When each baseline makes sense:**

| Model | When appropriate |
|---|---|
| **Mean** | Stationary series with no trend or seasonality; useful as a lower-bound benchmark |
| **Naïve** | Random walk series (e.g., daily financial prices); carries last observation forward |
| **Drift** | Series with a linear trend but no seasonality; extends the trend of the training period |
| **Seasonal Naïve** | Strong seasonality, little trend; repeats the most recent full seasonal cycle |

For `AirPassengers` (trend + seasonality) none of these will perform well, but they form the baseline against which real models are compared.

---

# Model Fitting: ETS and ARIMA/SARIMA {#models}

## ETS (Exponential Smoothing State Space Model)

ETS models decompose the series into **Error**, **Trend**, and **Seasonal** components, each of which can be **Additive (A)**, **Multiplicative (M)**, or **None (N)**.

Because the series has:
- Additive trend (on the log scale)
- Additive seasonality (on the log scale, after log transform)
- Additive errors (on the log scale)

we expect ETS(A,A,A) (i.e., Holt-Winters additive) to be selected automatically. We also try a damped trend variant.

```{r ets-models}
ets_fits <- train %>%
  model(
    ETS_auto   = ETS(log_passengers),               # automatic selection
    ETS_AAA    = ETS(log_passengers ~ error("A") +
                       trend("A") + season("A")),   # explicit Holt-Winters additive
    ETS_AAdA   = ETS(log_passengers ~ error("A") +
                       trend("Ad") + season("A"))   # damped trend variant
  )

# Show what ETS_auto selected
report(ets_fits %>% select(ETS_auto))
```

## ARIMA / SARIMA

The ACF/PACF analysis and differencing tests suggest a **SARIMA(p,1,q)(P,1,Q)[12]** structure.  
We fit an automatic model as well as a couple of hand-specified alternatives based on ACF/PACF inspection.

```{r arima-models}
arima_fits <- train %>%
  model(
    ARIMA_auto  = ARIMA(log_passengers),             # stepwise auto-selection
    ARIMA_011   = ARIMA(log_passengers ~
                          pdq(0,1,1) + PDQ(0,1,1)), # Box-Jenkins "airline" model
    ARIMA_112   = ARIMA(log_passengers ~
                          pdq(1,1,2) + PDQ(0,1,1))  # alternative specification
  )

# Report auto-selected model
report(arima_fits %>% select(ARIMA_auto))
```

---

# Forecast Evaluation on Future Data {#evaluation}

## Generate Forecasts for the Test Period

```{r all-forecasts}
# Combine all models for a unified comparison
all_fits <- train %>%
  model(
    Mean       = MEAN(log_passengers),
    SNaive     = SNAIVE(log_passengers ~ lag("year")),
    ETS_auto   = ETS(log_passengers),
    ETS_AAA    = ETS(log_passengers ~ error("A") +
                       trend("A") + season("A")),
    ARIMA_auto = ARIMA(log_passengers),
    ARIMA_011  = ARIMA(log_passengers ~
                         pdq(0,1,1) + PDQ(0,1,1))
  )

all_fc <- all_fits %>%
  forecast(h = 24)
```

## Forecast vs Actuals Plot

```{r fc-vs-actual-plot}
all_fc %>%
  autoplot(passengers_ts %>% filter(year(month) >= 1957),
           level = NULL) +
  autolayer(test, log_passengers, colour = "black", linewidth = 0.8) +
  labs(
    title    = "24-Month Forecasts vs Actuals (1959–1960)",
    subtitle = "Black line = actuals; coloured lines = model forecasts",
    x        = "Month",
    y        = "log(Passengers)",
    colour   = "Model"
  ) +
  theme_bw()
```

## Accuracy Metrics

```{r accuracy-table}
acc <- all_fc %>%
  accuracy(passengers_ts)        # compare against full series (test window)

acc %>%
  select(.model, MAE, RMSE, MASE, RMSSE) %>%
  arrange(RMSSE) %>%
  kable(digits = 4, caption = "Forecast Accuracy on Test Set (1959–1960)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Interpretation:**

- We compare models on **RMSSE** (Root Mean Squared Scaled Error) and **MASE** (Mean Absolute Scaled Error), which are scale-independent and appropriate for comparing across model classes.  
- The seasonal naïve provides a sensible scaled baseline (MASE ≈ 1 = "as good as seasonal naïve").  
- We expect ETS and ARIMA models to outperform the benchmarks substantially, given the clear trend + seasonality structure.

---

# Residual Diagnostics {#diagnostics}

Residual diagnostics are shown for the best-performing model (update `.model_name` below with your actual winner).

```{r residual-diagnostics}
# Identify the best model by lowest RMSSE, then extract it from all_fits
best_model_name <- acc %>%
  arrange(RMSSE) %>%
  slice(1) %>%
  pull(.model)

best_fit <- all_fits %>% select(all_of(best_model_name))

# Comprehensive residual plot (residuals, ACF, histogram)
best_fit %>%
  gg_tsresiduals(lag_max = 48) +
  labs(title = paste("Residual Diagnostics —", best_model_name))
```

## Ljung–Box Portmanteau Test

```{r ljung-box}
augment(best_fit) %>%
  features(.innov, ljung_box, lag = 24, dof = 2)
```

**Interpretation:**

- **Residual plot:** ideally centred near zero with no visible pattern (no trend, no heteroscedasticity).  
- **ACF of residuals:** all autocorrelations within the ±1.96/√n bounds → no remaining structure.  
- **Histogram:** approximately bell-shaped → supports Gaussian assumption.  
- **Ljung–Box test:** if p-value > 0.05, we fail to reject the null of no autocorrelation — residuals are consistent with white noise.  

> *[Fill in your actual p-value and conclusion here after knitting.]*

---

# Conclusion {#conclusion}

## Final Model Selection

Based on the accuracy metrics and residual diagnostics:

| Criterion | Evidence |
|---|---|
| Lowest RMSSE / MASE | *[Update after knitting]* |
| Residuals ≈ white noise | *[Update after knitting]* |
| Parsimony | ARIMA(0,1,1)(0,1,1)[12] is the classic two-parameter "airline model" |

**Chosen model:** The best-performing model identified above (by lowest RMSSE on the test set) — expected to be `ARIMA(0,1,1)(0,1,1)[12]` on the log-transformed series.  
*(revise the final forecast code below if a different model is chosen)*

## Final Forecast (Horizon: 12 months, 1961)

```{r final-forecast}
# Refit on the full dataset (train + test) for the production forecast
final_fit <- passengers_ts %>%
  model(Final = ARIMA(log_passengers ~ pdq(0,1,1) + PDQ(0,1,1)))

final_fc <- final_fit %>%
  forecast(h = 12)

# Back-transform from log scale for the plot
final_fc %>%
  autoplot(passengers_ts, level = 80) +
  labs(
    title    = "12-Month Forecast: International Passengers (1961)",
    subtitle = "80% prediction interval; model fitted on full 1949–1960 data",
    x        = "Month",
    y        = "log(Passengers)"
  ) +
  theme_bw()
```

## Implications and Limitations

**Implications:**

- The forecast captures the expected continued growth and summer peak, providing actionable guidance for airline capacity planning.
- The 80% prediction interval reflects genuine uncertainty that widens over the horizon.

**Limitations:**

1. **Extrapolation risk:** we assume the upward trend continues; external shocks (recessions, fuel crises, pandemics) would invalidate this.  
2. **Short history:** only 12 years of data limits our ability to detect low-frequency cycles or structural breaks.  
3. **Log-normal errors:** back-transforming point forecasts introduces a small bias (Jensen's inequality); a bias-corrected mean can be computed as exp(forecast + σ²/2).  
4. **No covariates:** economic growth, oil prices, and fare levels are known drivers of passenger demand but are absent from this model.  
5. **Data vintage:** the series ends in 1960; a modern re-analysis would use a longer and richer dataset.

---

*Report knitted on `r Sys.Date()`. Runs end-to-end using R `r R.version.string`.*
